---
title: "Basic principles"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Basic principles}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, results = 'hide'}
suppressPackageStartupMessages(library(SCDB))
options("SCDB.log_path" = tempdir())
```

The basic principle of the SCDB package is to enable the user to easily implement and maintain a database of time-versioned data.

In practice, this is done by labeling each record in the data with three additional fields:

* checksum: An md5 hash of information stored the record
* from\_ts: the timestamp at which the record was introduced to the dataset
* until\_ts: the timestamp at which the record was removed from the dataset

This strategy of time versioning is often called "type 2" history.

Note that identical records may be removed and introduced more than once; for example, in a table of names and addresses, a person may change their address (or name) back to a previous value.

The SCDB package provides the function `update_snapshot` to handle the insertion and deactivation of records using this strategy.
It further includes several functions to improve the Quality of life for working with database data.

A simple example is shown below:

```{r example_conn}
# First we connect to our database.
# If this is, e.g., a PostgreSQL database already running on the machine, connection
# can be done after the configuration of a .pgpass file.
# For this example, we use an on-disk SQLite db to showcase.
conn <- get_connection(drv = RSQLite::SQLite())
# NOTE: Had the PostgreSQL DB been configured, we would not need to pass any args
# to get_connection()
```

```{r example_data}
# Our example data is mtcars with rownames converted to a column and only the first hp
# column of mtcars
example_data <- dplyr::transmute(mtcars, car = rownames(mtcars), hp)
# If the data does not already live on the remote, we must transfer it
example_data <- dplyr::copy_to(conn, example_data, overwrite = TRUE)

```

```{r example_1}
# In this example, we imagine that on day 1, in this case 2020-01-01 11:00:00, our data
# known to us is the first 3 records of mtcars
data <- head(example_data, 3)
```

```{r update_snapshot_1, eval = FALSE}
# We then store these data in the database using update_snapshot
update_snapshot(.data = data,
                conn = conn,
                db_table = "mtcars", # the name of the DB table to store the data in
                timestamp = as.POSIXct("2020-01-01 11:00:00"))
```
```{r update_snapshot_1_hidden, results = 'hide', echo = FALSE}
# We then store these data in the database using update_snapshot
invisible(capture.output({  # NOTE: update_snapshot does some printing to terminal
  update_snapshot(.data = data,
                  conn = conn,
                  db_table = "mtcars", # the name of the DB table to store the data in
                  timestamp = as.POSIXct("2020-01-01 11:00:00"))
}))
```

```{r example_1_results}
# We can access our data using the `get_table` function
print(get_table(conn, "mtcars"))

# And we can see the time-keeping if we set `include_slice_info = TRUE`
print(get_table(conn, "mtcars", include_slice_info = TRUE))
```

```{r example_2}
# Let's say that the next day, our data set is now the first 5 of our example data
data <- head(example_data, 5)
```

```{r update_snapshot_2, eval = FALSE}
# We then store these data in the database using update_snapshot
update_snapshot(.data = data,
                conn = conn,
                db_table = "mtcars", # the name of the DB table to store the data in
                timestamp = as.POSIXct("2020-01-02 12:00:00"))
```
```{r update_snapshot_2_hidden, results = 'hide', echo = FALSE}
# We then store these data in the database using update_snapshot
invisible(capture.output({
  update_snapshot(.data = data,
                  conn = conn,
                  db_table = "mtcars", # the name of the DB table to store the data in
                  timestamp = as.POSIXct("2020-01-02 12:00:00"))
}))
```

```{r example_2_results}

# We again use the `get_table` function and see the latest available data
print(get_table(conn, "mtcars"))

# And we can see the time-keeping if we set `include_slice_info = TRUE`
print(get_table(conn, "mtcars", include_slice_info = TRUE))

# Since our data is time-versioned, we can recover the data from the day before
print(get_table(conn, "mtcars", slice_ts = "2020-01-01 11:00:00"))
```

```{r example_3}
# On day 3, we imagine that we have the same 5 records, but one of them is altered
data <- head(example_data, 5) |>
  dplyr::mutate(hp = ifelse(dplyr::row_number() == 1, hp / 2, hp))
```

```{r update_snapshot_3, eval = FALSE}
# We then store these data in the database using update_snapshot
update_snapshot(.data = data,
                conn = conn,
                db_table = "mtcars", # the name of the DB table to store the data in
                timestamp = as.POSIXct("2020-01-03 10:00:00"))
```
```{r update_snapshot_3_hidden, results = "hide", echo = FALSE}
# We then store these data in the database using update_snapshot
invisible(capture.output({
  update_snapshot(.data = data,
                  conn = conn,
                  db_table = "mtcars", # the name of the DB table to store the data in
                  timestamp = as.POSIXct("2020-01-03 10:00:00"))
}))
```

```{r example_3_results}
# We can again access our data using the `get_table` function and see that the currently
# available data (with the changed hp value for Mazda RX4)
print(get_table(conn, "mtcars"))


# When `slice_ts` is set to `NULL`, the full history of the table is returned
print(get_table(conn, "mtcars", slice_ts = NULL))

# Setting include_slice_info = TRUE also returns checksum, from_ts and until_ts.
# This is most useful when viewing data from a specific point in time
print(get_table(conn, "mtcars", slice_ts = "2020-01-03 06:30:00",
                include_slice_info = TRUE))
```
