---
title: "SCDB: Benchmarks"
output: rmarkdown::html_vignette
bibliography: "references.bib"
vignette: >
  %\VignetteIndexEntry{SCDB: Benchmarks}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

# NOTE:
# To re-run the benchmarks, run the "benchmark" workflow on GitHub
```

```{r}
library(magrittr)
```

To showcase the performance of `SCDB` on different database backends, we include this vignette that summarises a simple
benchmark:
A sample dataset is created based on the `datasets::iris` dataset. This data is repeated 10 times and given a unique ID
(the row number of the data).

This data forms the basis for three "snapshots" used in the benchmarks:

1) The data as described above.

2) As 1. but where any `Sepal.Length` below the median is halved.

3) As 2. but where any `Sepal.width` below the median is halved.

The benchmark function uses three consecutive calls to `update_snapshot()` to create the table with first snapshot and
then update it to the second and third snapshot. Finally, the table is deleted.

The performance of this benchmark function is timed with the `{microbenchmark}` package using 10 replicates.
All benchmarks are run on the same machine.

The results of the benchmark are shown graphically below (mean and standard deviation), where we compare the current
development version of `SCDB` with the current CRAN version.

```{r benchmark_preprocessing, echo = FALSE, eval = requireNamespace("here")}
benchmark_location <- c(
  system.file("extdata", "benchmarks.rds", package = "SCDB"),
  here::here("inst", "extdata", "benchmarks.rds")
) %>%
  purrr::discard(~ identical(., "")) %>%
  purrr::pluck(1)

benchmarks <- readRDS(benchmark_location)

# Determine if the SHA is on main
sha <- benchmarks %>%
  dplyr::distinct(version) %>%
  dplyr::filter(!(version %in% c("CRAN", "main", "branch"))) %>%
  dplyr::pull("version")

# Check local git history
on_main <- tryCatch({
  system(glue::glue("git branch main --contains {sha}"), intern = TRUE) %>%
    stringr::str_detect(stringr::fixed("main")) %>%
    isTRUE()
}, warning = function(w) {
  # If on GitHub, git is not installed and we assume TRUE.
  # This will render the vignette as it will look once merged onto main.
  return(identical(Sys.getenv("CI"), "true"))
})

# If the SHA has been merged, use as the "main" version and remove the other, older, main version
if (on_main) {
  benchmarks <- benchmarks %>%
    dplyr::filter(version != "main") %>%
    dplyr::mutate(version = dplyr::if_else(.data$version == "CRAN", "CRAN", "development"))
}

# Mean and standard deviation (see ggplot2::mean_se())
mean_sd <- function(x) {
  mu <- mean(x)
  sd <- sd(x)
  data.frame(y = mu, ymin = mu - sd, ymax = mu + sd)
}
```

```{r benchmark_1, echo = FALSE, eval = requireNamespace("here")}
# Use data for benchmark 1
benchmark_1 <- benchmarks %>%
  dplyr::filter(!stringr::str_ends(.data$benchmark_function, stringr::fixed("complexity")))

# Add note slow backends
slow_backends <- benchmark_1 %>%
  dplyr::distinct(.data$database, .data$n) %>%
  dplyr::filter(.data$n < max(.data$n)) %>%
  dplyr::pull("database")

benchmark_1 <- benchmark_1 %>%
  dplyr::mutate("database" = paste0(database, ifelse(database %in% slow_backends, "*", "")))

# Insert newline into database name to improve rendering of figures
labeller <- ggplot2::as_labeller(function(l) stringr::str_replace_all(l, stringr::fixed(" v"), "\nv"))



g <- ggplot2::ggplot(
  benchmark_1,
  ggplot2::aes(x = version, y = time / 1e9)
) +
  ggplot2::stat_summary(fun.data = mean_sd, geom = "pointrange", size = 0.5, linewidth = 1) +
  ggplot2::facet_grid(
    rows = ggplot2::vars(benchmark_function),
    cols = ggplot2::vars(database),
    labeller = labeller
  ) +
  ggplot2::labs(x = "Codebase version", y = "Time (s)")

if (length(slow_backends) > 1) {
  g <- g + ggplot2::labs(caption = "* IMPORTANT: Benchmark data halved for this backend!")
}

g
```


We include another benchmark to highlight the complexity scaling of the `update_snapshot() ` with the size of the input
data. The datasets are similar to the first benchmark, but the number of repeats is varied to see the impact of
increasing data size. The benchmarks are run from a "clean" state, where the target_table does not exists. The benchmark
measures both the time to create the table and to remove it again afterwards (to restore the clean state).

The performance of this benchmark function is timed with the `{microbenchmark}` package using 5 replicates.
All benchmarks are run on the same machine.

The results of the benchmark are shown graphically below (mean and standard deviation) and with linear scaling (dotted
line), where we compare the current development version of `SCDB` with the current CRAN version.

NOTE: There are reports of a superlinear complexity for very large data sets. If you experience such problems, consider
batching the updates via the `filters` argument.

```{r benchmark_2, echo = FALSE, eval = requireNamespace("here")}
# Use data for benchmark 2
benchmark_2 <- benchmarks %>%
  dplyr::filter(stringr::str_ends(.data$benchmark_function, stringr::fixed("complexity"))) %>%
  dplyr::mutate("benchmark_function" = stringr::str_remove_all(benchmark_function, stringr::fixed("- complexity")))

ggplot2::ggplot(
  benchmark_2,
  ggplot2::aes(x = n * nrow(iris) / 1e3, y = time / 1e9, color = version)
) +
  ggplot2::stat_summary(fun.data = mean_sd, geom = "pointrange", size = 0.5, linewidth = 1) +
  ggplot2::geom_smooth(method = "lm", formula = y ~ x, se = FALSE, linetype = 3) +
  ggplot2::facet_grid(
    rows = ggplot2::vars(benchmark_function),
    cols = ggplot2::vars(database),
    labeller = labeller
  ) +
  ggplot2::labs(x = "Data size (1,000 rows)", y = "Time (s)", color = "Codebase version") +
  ggplot2::theme(panel.spacing = grid::unit(1, "lines"), legend.position = "bottom")
```
