---
title: "SCDB: Benchmarks"
output: rmarkdown::html_vignette
bibliography: "references.bib"
vignette: >
  %\VignetteIndexEntry{SCDB: Benchmarks}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

To showcase the performance of SCDB on different database backends, we include this vignette that summarises a simple
benchmark:
A sample dataset is created based on the `datasets::iris` dataset. This data is repeated 10 times and given a unique ID
(the row number of the data).

This data forms the basis for three "snapshots" used in the benchmarks:

1) The data as described above.

2) As 1. but where any `Sepal.Length` below the median is halved.

3) As 2. but where any `Sepal.width` below the median is halved.

The benchmark function uses three consecutive calls to `update_snapshot()` to create the table with first snapshot and
then update it to the second and third snapshot. Finally, the table is deleted.

The performance of this benchmark function is timed with the `microbenchmark` package using 10 replicates.
All benchmarks are run on the same machine.

The results of the benchmark are shown graphically below (mean and standard deviation), where we compare the current
development version of SCDB with the current CRAN version.

```{r, echo = FALSE, eval = requireNamespace("here")}
benchmark_location <- c(
  system.file("extdata", "benchmarks.rds", package = "SCDB"),
  here::here("inst", "extdata", "benchmarks.rds")
) |>
  purrr::discard(~ identical(., "")) |>
  purrr::pluck(1)

benchmarks <- readRDS(benchmark_location)

# Determine if the SHA is on main
sha <- benchmarks |>
  dplyr::distinct(version) |>
  dplyr::filter(!(version %in% c("CRAN", "main"))) |>
  dplyr::pull("version")

# Check local git history
on_main <- tryCatch({
  system(glue::glue("git branch main --contains {sha}"), intern = TRUE) |>
    stringr::str_detect(stringr::fixed("main")) |>
    isTRUE()
}, warning = function(w) {
  # If on GitHub, git is not installed and we assume TRUE.
  # This will render the vignette as it will look once merged onto main.
  return(identical(Sys.getenv("CI"), "true"))
})

# If the SHA has been merged, use as the "main" version and remove the other, older, main version
if (on_main) {
  benchmarks <- benchmarks |>
    dplyr::filter(version != "main") |>
    dplyr::mutate(version = dplyr::if_else(.data$version == "CRAN", "CRAN", "development"))
}

# Add note slow backends
slow_backends <- benchmarks |>
  dplyr::distinct(.data$database, .data$n) |>
  dplyr::filter(.data$n < max(.data$n)) |>
  dplyr::pull("database")

benchmarks <- benchmarks |>
  dplyr::mutate("database" = paste0(database, ifelse(database %in% slow_backends, "*", "")))


# Mean and standard deviation (see ggplot2::mean_se())
mean_sd <- function(x) {
  mu <- mean(x)
  sd <- sd(x)
  data.frame(y = mu, ymin = mu - sd, ymax = mu + sd)
}

g <- ggplot2::ggplot(
  benchmarks,
  ggplot2::aes(x = version, y = time / 1e9)
) +
  ggplot2::stat_summary(fun.data = mean_sd, geom = "pointrange", size = 0.5, linewidth = 1) +
  ggplot2::facet_grid(rows = ggplot2::vars(benchmark_function), cols = ggplot2::vars(database)) +
  ggplot2::labs(x = "Codebase version", y = "Time (s)")

if (length(slow_backends) > 1) {
  g <- g + ggplot2::labs(caption = "* IMPORTANT: Benchmark data halved for this backend!")
}

g
```
